---
title: "TP1 - Data Exploration"
author: "Arthur Capet"
date: "2024-03-17"
pdf_document:
    toc: yes
    toc_depth : 3
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(readxl)
```


# Loading Packages and Functions
Load the required packages
(vegan must be loaded after ade4 to avoid some conflicts)

```{r}

library(readxl)

# library(vegan)        
# 
# source("../Functions/panelutils.R")
# source("../Functions/coldiss.R")
# source("../Functions/evplot.R")
# source("../Functions/hcoplot.R")
# source("../Functions/Answers.R")

# library(lubridate)
# library(gdata)
# library(ggplot2)
# require(reshape2)
# require(plyr)
# 
# library(ade4)
# library(cluster)
# library(RColorBrewer)

# Package not available currently
## library(labdsv)
## library(kableExtra)
```


# Part 1 : Basic R.

## Variables
In R we can assign value to variables, and use them afterwards for computation.

```{r}
a <- 4
print(4+a)
```

*Variables* can have different types : 
  - "numeric" is a number (real).
  - "chr" is a character string.

```{r}
# We can also include comment lines in code blocs.
print("The type and value of 'a' are :")
typeof(a)
print(a)
```


```{r}
# Let's define a "character string" variables
b <- 'Hello'
print("The type and value  of 'b' is :")
typeof(b)
print(b)
```


```{r}
print("The type and value  of 'TRUE' is :")
typeof(TRUE)
print(TRUE)
```


```{r}
myflag <- 1+1==2
print("The type and value  of myflag is :")
typeof(myflag)
print(myflag)
```

Those are variables that contain only a single element.
Sometimes, it is usefull to build vector variables that contains several elements.

```{r}
A <- c(1,2,3)
print(A)
```

Vectors can be built for variables of any type.
```{r}
B <- c('Hello','World')
B
```
```{r}
C <- A < 3
print(C)
```
Of course, methods exists to build vectors in one go : 
```{r}
print(seq (from = 1, to = 8, by = 2))
print(1:4)
print(5:-1)
```


## Functions

In general, we use functions apply the same operation to different arguments.

```{r}
squaremynumber <- function(number){
  return(number*number)
}

print("The type and value  of 'squaremynumber' is :")
typeof(squaremynumber)

print (squaremynumber(2))

print (squaremynumber(5))

print (squaremynumber(A))
```


# Part 2 : Data load and manipulation 

We received monitoring data from the Lake Nokoué.
These have been stored in an '.xls' file. Our first objective is to read these data and import them in an interactive R session.

```{r}
filename <- system.file("data-raw", "nokoue.xlsx", package="PhytoNokoueTPR")
print(filename)
```


```{r}
dfmai <- read_excel(filename,"Mai")
dfmai
```

The `dfmai` variable now contains a **Tibble Dataframe** (we'll just use the word "**dataFrame**").
This is a very convenient variable type for data analysis. It contains different columns ("Variables") and rows ("Observations").
The nice thing with that type is that different columns can have different types (off course all observations in a given column has to have the same type).

The content of a given column can be accessed simply : 
```{r}
print(dfmai$T)
```

.. and there are plenty of command to manipulate them, for instance for subsetting,  

```{r}
subset(dfmai, dfmai$T>29.5)
```

What we need to do now is :
  - Ensure that observations have the correct type. 
  - Gather all months in a single large data frame. 
  - Build a descriptor dataframe with units and definitions. 
  
## Observation type

The columns in our data frame are of different nature. We need to distinguish : 

  - Coordinates variables : Those are not analysed **per se**, but are used as a reference basis to analyze other variables. 
      - Sites : The name of each sampling site (ie. Spatial Coordinate)
      - Mois : The month of each sampling site (ie. Temporal Coordinate)
      - X, Y : Geospatial references, i.e. to produce maps. 
  - Environmental variables : 
      
      
