---
pdf_document:
  toc: true
  toc_depth: 3
author: "Arthur Capet"
date: "2024-03-17"
output: pdf_document
title: "Workshop PhytoNokoué - TP1 - Data Loading"
---

```{r setup, include=FALSE}
# Loading required packages
knitr::opts_chunk$set(echo = TRUE, results='hide')
library(readxl)
library(ggplot2)
library(reshape2)
library(dplyr)
```


# JOUR 1

## Part 1 : Basic R.

### Variables

In R we can assign value to variables, and use them afterwards for computation.

```{r}
a <- 5
print(4+a)
```

*Variables* can have different types : 
  - "numeric" is a number (real).
  - "chr" is a character string.

```{r}
# We can also include comment lines in code blocs.
print("The type and value of 'a' are :")
typeof(a)
print(a)
```


```{r}
# Let's define a "character string" variables
b <- 'Hello'
print("The type and value of 'b' is :")
typeof(b)
print(b)
```


```{r}
print("The type and value of 'TRUE' is :")
typeof(TRUE)
print(TRUE)
```


```{r}
myflag <- 1+1==2
print("The type and value of 'myflag' is :")
typeof(myflag)
print(myflag)
```

Those are variables that contain only a single element.
Sometimes, it is usefull to build vector variables that contains several elements.

```{r}
A <- c(5,8,13)
print(A)
print(A[2])
```

Vectors can be built for variables of any type.
```{r}
B <- c('Hello','World')
B
```
```{r}
C <- (A < 10)
print(C)
```

Of course, methods exists to build vectors implicitly : 
```{r}
print(seq (from = 1, to = 8, by = 2))
print(1:4)
print(5:-1)
```


### Functions

In general, we use functions to apply the same operation to different arguments.

```{r}
squaremynumber <- function(number){
  return(number*number)
}

print (squaremynumber(2))

print (squaremynumber(5))

print (squaremynumber(A))
```


## Part 2 : Data load and first manipulation 

We received monitoring data from the Lake Nokoué.
These have been stored in an '.xls' file. Our first objective is to read these data and import them in an interactive R session.

```{r}
filename <-"nokoue.xlsx"
print(filename)
```


```{r}
dfmai <- read_excel(filename,"Mai")
dfmai
```

The `dfmai` variable now contains a **Tibble Dataframe** (we'll just use the word "**dataFrame**").
This is a very convenient variable type for data analysis. It contains different columns ("Variables") and rows ("Observations").
The nice thing with that type is that different columns can have different types (off course all observations in a given column has to have the same type).

The content of a given column can be accessed simply : 
```{r}
print(dfmai$T)

print(dfmai[['T']])
```

When using a sinlge pair of bracket, the returned element is a one-column dataframe
```{r}
print(dfmai['T'])
```

.. and there are plenty of command to manipulate them, for instance for subseting,  

```{r}
subset(dfmai, dfmai$T>29.5)
```

What we need to do now is :
  - Build a descriptor dataframe with units and definitions for all variables 
  - Ensure that observations have the correct type. 
  - Gather all months in a single large data frame. 

### Variable Description Dataframe.

The columns in our data frame are of different nature. We need to distinguish : 

  - Coordinates variables: those are not analysed **per se**, but are used as a reference basis to analyze other variables. 
  - Environmental variables: those characterize the bio-chemo-physical environment, in our context we think of them as related to the "habitat". 
  - Species variables : those are counts of species (i.e. phytoplanctons).  

There is a dedicated sheet in the data file to describe the variables.

```{r}
dfvar <- read_excel(filename,"Units")
dfvar
```

To give an idea of why this is very useful, let's consider the following for loop : 

```{r}

for (v in colnames(dfmai)){
  varname <- v
  local_dfvar <- subset(dfvar, dfvar$Parameters==v)
  vartype <- local_dfvar$Type
  varunit <- local_dfvar$Units
  varname <- local_dfvar$Longname
  print(paste(v," is a ", vartype, "variable, expressed in ", varunit, " and representing", varname  ))
  
}

```

# JOUR 2
  
### Observation type

But Species counts should un fact be given as integers !! 

We can fo this manually, column by column :   
```{r}
dfmai$EUGL <-as.integer(dfmai$EUGL)
dfmai
```

But this is boring, so let us use our variable description dataframe instead !

```{r}
for (v in subset(dfvar, dfvar$Type=="Species")$Parameters){
  print(v)
  dfmai[[v]] <-as.integer(dfmai[[v]]) 
}

dfmai
```
      
### Combining dataframe for the different months. 

We have one sheet per month. It will be much more convenient to have all data in a single data frame. 
Since we have to apply the conversion to integer, we can make a function to load a monthly data frame. 

```{r}
# A function to load monthly data frames
load_monthly_df <- function(monthstr){
  # Parameters
  #   - monthsr - an character string corresponding to one of the sheets of our .xlsx file.
  # Value
  #   - A data frame with data for one month.
  
  # Read the corresponding sheet
  df <- read_excel(filename,monthstr)
  
  # Loop on Species parameters to convert towards integers.
  for (v in subset(dfvar, dfvar$Type=="Species")$Parameters){
    df[[v]] <-as.integer(df[[v]]) 
  }
  
  # The command to return the result of the function.
  return(df)
}
```

This is just to check that the function works.

```{r}
df <- load_monthly_df('Fevrier')

df
```

Now we will use this function in a loop, to build *a list of dataframes*, i.e. a single variable object that includes all the dataframe.

```{r}
listofdf <- list()

monthnames <- c('Janvier','Fevrier','Mars','Avril','Mai','Juin','Juillet','Aout','Septembre','Octobre','Novembre','Decembre')

for (i in 1:12){
  print(monthnames[i])
  df <- load_monthly_df(monthnames[i])
  listofdf[[i]] <- df
}
```

Now we have a list of dataframes, but we still need to gather them in a single dataframe. This operation is called concatenate, and can be achieved with the function `rbind`, where 'r' stands for 'rows' (the different dataframes are to be considered as group of *rows* of a larger dataframe).

```{r}
dfdata <- do.call(rbind, listofdf)
```

Our dataset is ready .. or almost.

## First plots (using ggplot)

Ggplot is a very powerful tool to explore datasets. It as a very particular syntax. The step to consider are : 

1. Which is the dataframe I want to use for my plot.
2. What variable should be used for what aspects (or "Aestethic") of the graph ( think, on what shold be based the x-axis coordinate, y-axis coordinate, the color, the size of the points, etc ...)
3. What kind of plot do i want to produce ? Points? Lines ? Bars ? 

Here is a simple example, where we use the dataframe `dfdata`, to build a scatterplot (points) with showing temperature, as a function of the months.

```{r}
ggplot(dfdata, aes(x=Mois, y = T))+geom_point()
```

There is an issue, because the variable `Mois` is just a series of character strings, ggplot doesn't know in which order they should appear.
We can clarify this, byt defining this column as a set of 'factors'. Factors are discrete categorical variables, but they can be defined to be ordered.


```{r}
dfdata$Mois <- factor(dfdata$Mois, ordered=TRUE, levels=monthnames)

ggplot(dfdata, aes(x=Mois, y = T))+geom_point()
```


If we want to show stations in colors, we just have to do :       
```{r}
ggplot(dfdata, aes(x=Mois, y = T, color = Sites))+geom_point()
```
      

Then, of course, there is plenty of way to modify graphical apsects of the figure :
```{r}
ggplot(dfdata, aes(x=Mois, y = T, color = Sites))+
  geom_point()+
  theme_bw()+ theme(axis.text.x = element_text(angle = 60, vjust = 0.5, hjust=1))
```
To show sme cool feature of ggplot, we need month as a numeric variables. Let's add it to the data.frame.
Because I'm lay, I'll just the integer underlying the factor format we have given to `dfdata$Mois`.

```{r}
dfdata$Monthnumber <- as.integer(dfdata$Mois)
```

      
```{r}
ggplot(dfdata, aes(x=Monthnumber, y = T))+
  geom_smooth()+
  theme_bw()+ theme(axis.text.x = element_text(angle = 60, vjust = 0.5, hjust=1))
```

      
```{r}
ggplot(dfdata, aes(x=Mois, y = T))+
  geom_boxplot()+
  theme_bw()+ theme(axis.text.x = element_text(angle = 60, vjust = 0.5, hjust=1))

```


      
```{r}
ggplot(dfdata, aes(x=Mois, y = T))+
  geom_boxplot()+
  geom_point(aes(color = Sites))+
  theme_bw()+ theme(axis.text.x = element_text(angle = 60, vjust = 0.5, hjust=1))
```


### Exercices GGplot

Faire des boxplots de la Salinité en fonction des Stations

```{r}

```



Build a scatter plot of your favorite chemical variable as a function of Salinity.

```{r}

```


## dplyr package, to analyse dataframe

Split - Apply - Combine 

```{r}
dfdata %>% group_by(Sites) %>% summarise(MeanT=mean(T))
```


## Long and Wide format      
In general, it is easier to work in ggplot with 'long' table format, rather than 'wide' table format. 
We can pass from 'wide' to 'long' using the command `melt`.
      
```{r}
ldf <- melt(dfdata, id.vars = subset(dfvar, dfvar$Type=='Coordinate')$Parameters )
ldf
```



```{r}
ldf %>% group_by(Sites, variable) %>% summarise(Mean=mean(value))
```

Exercice : use dply to identify median value of all variables per month.
```{r}

```

## GG facet plots
Now, we can use a bit more advanced ggplot nomenclature to build more complex plots.

```{r, fig.height=15, fig.width=10}
envdf <- subset(ldf, ldf$variable %in% subset(dfvar, dfvar$Type == 'Environment')$Parameters)

ggplot(envdf, aes(x=Mois, y = value))+
  geom_point(aes( color=Sites))+
  geom_boxplot(alpha=.7)+
  facet_wrap(vars(variable), scales = "free_y", ncol=3)+
  theme_bw()+ theme(axis.text.x = element_text(angle = 60, vjust = 0.5, hjust=1))
```

Find a way to represent the spatiual variability for a given variable

```{r, fig.height=15, fig.width=10}

```

