---
title: "Phytoplankton inventory and relationships with environmental factors in Nokoué lake in Bénin (West Africa)"
author: "Firmin M. ADANDEDJI1, 2, Arthur CAPET3, Héloise LAVIGNE3, Anne Julie ROCHETTE3, Daouda MAMA1,2"
date: "2024-05-28"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(readxl)
library(ggplot2)
library(reshape2)
library(dplyr)
library(lubridate)


library(ade4)
library(vegan)        
library(gclus)        

library(cluster)
library(RColorBrewer)
library(gridExtra)

source('./coldiss.R')
source('./hcoplot.R')
source('./cleanplot.PCA.R')
source('./panelutils.R')
```


```{r}
filename <-"./Data_phytonokoue_new.xlsx"
df <-  read_excel(filename,  col_types = c(c("text", "numeric", "numeric", "text", "text", "date"), rep('numeric', 24)))
# Arranging date format
df$datetime <- dmy_hms(paste(df$Date,format(df$Heure, "%H:%M:%S")))
df <- df %>% select(-Date, -Heure, -Month)

df
```

```{r}
dfvar <- read_excel(filename,"Units")
dfvar <- dfvar %>% add_row(Parameters = "datetime", Unit = "-", Type = "coord") %>%
  add_row(Parameters = "X", Unit = "-", Type = "coord") %>%
  add_row(Parameters = "Y", Unit = "-", Type = "coord") %>% 
  add_row(Parameters = "Sites", Unit = "-", Type = "coord")

dfvar
```


```{r}
ldf <- melt(df, id.vars = subset(dfvar, dfvar$Type=='coord')$Parameters )
```


```{r, fig.height=10, fig.width=8}

envdf <- subset(ldf, ldf$variable %in% subset(dfvar, dfvar$Type == 'env')$Parameters)


envdf$Mois <- as.numeric(month(envdf$datetime))#factor(month(envdf$datetime), ordered=TRUE, levels=month.abb)


ggplot(envdf, aes(x=Mois, y = value))+
  geom_point(aes( color=Sites))+
  geom_boxplot(aes(group=Mois),alpha=.7)+
  facet_wrap(vars(variable), scales = "free_y", ncol=4)+
  theme_bw()+theme(legend.position='bottom')+ 
  theme(axis.text.x = element_text(angle = 60, vjust = 0.5, hjust=1))+
  guides(color=guide_legend(nrow=2,byrow=TRUE))+
  xlab("Months")+ylab('[log(ind/mL)]')+scale_x_continuous(breaks = 1:12, labels = month.abb)

```


```{r, fig.height=8, fig.width=10}
spedf <- subset(ldf, ldf$variable %in% subset(dfvar, dfvar$Type == 'spe')$Parameters)
spedf$Mois <- as.numeric(month(spedf$datetime))
```

```{r, fig.height=8, fig.width=10}
ggplot(spedf, aes(x=month(datetime), y = log(value+1)))+
  geom_point(aes( color=Sites))+
  geom_boxplot(aes(group=month(datetime)),alpha=.7)+
  facet_wrap(vars(variable), ncol=4)+
  theme_bw()+ theme(axis.text.x = element_text(angle = 60, vjust = 0.5, hjust=1))+
  xlab("Months")+ylab('[log(ind/mL)]')+scale_x_continuous(breaks = 1:12, labels = month.abb)
```


```{r, fig.height=8, fig.width=10}
ggplot(spedf, aes(x=Sites, y = log(value+1)))+
  geom_point(aes(color=Sites))+
  geom_boxplot(aes(group=Sites),alpha=.7)+
  facet_wrap(vars(variable), ncol=2, scale='free_y')+
  theme_bw()+ theme(axis.text.x = element_text(angle = 60, vjust = 0.5, hjust=1))+
  xlab("Months")+ylab('[log(ind/mL)]')#+scale_x_continuous(breaks = 1:12, labels = month.abb)
```

## Maps


```{r}
library(ggmap)
bmap <- get_stadiamap(bbox = c(left = 2.33, right = 2.57, bottom = 6.35, top=6.53), maptype = "stamen_terrain", zoom = 12, where='.')
```

```{r}
monthnames <-  month.abb
names(monthnames) <- seq(1:12)
```


## Mapping

```{r, fig.width=12}
mapmyenv<- function (eee){
  p <- ggmap(bmap)+
  geom_point(subset(envdf, variable==eee),  mapping = aes(x = X, y = Y, color=value))+facet_wrap('Mois',labeller=as_labeller(monthnames))+scale_color_viridis_c(name= paste(subset(dfvar, Parameters==eee)$Title,subset(dfvar, Parameters==eee)$Unit, sep='\n' ))
  return(p)
}

mapmyspe<- function (eee){
  p <- ggmap(bmap)+
    geom_point(subset(spedf, variable==eee),  mapping = aes(x = X, y = Y,color=log(value+1)))+
    facet_wrap('Mois',labeller=as_labeller(monthnames))+
    scale_color_viridis_c(name= paste(subset(dfvar, Parameters==eee)$Title,"[log(ind/mL)]", sep='\n' ))
  return(p)
}
```


```{r}
if (TRUE){
  # Saves monthly maps in PNG file for every environmental variables
  for (eee in subset(dfvar, Type=="env")$Parameters){
    png(paste0("./",eee,"_map.png"), width = 8, height = 5, unit="in", res=200)
    p <- mapmyenv(eee)
    print(p)
    dev.off()
  }
  
  # Saves monthly maps in PNG file for every species variables
  for (eee in subset(dfvar, Type=="spe")$Parameters){
    png(paste0("./",eee,"_map.png"), width = 8, height = 5, unit="in", res=200)
    p <- mapmyspe(eee)
    print(p)
    dev.off()
  }
}
```


```{r, eval=FALSE}

for (eee in subset(dfvar, Type=="env")$Parameters){
  p <- mapmyenv(eee)
  print(p)
}
```



```{r}

# Saves monthly maps in PNG file for every species variables
for (eee in subset(dfvar, Type=="spe")$Parameters){
  p <- mapmyspe(eee)
  print(p)
}
```


# Species 

Some summary stats for the plankton counts : 
```{r}

spesum <- spedf %>% 
  filter(!is.na(value)) %>%
  group_by(variable, month(datetime)) %>% 
  summarize(
    spotstation=sum(value>0),
    max.count=max(value),
    mean.count=mean(value),
    mean.count.logp1=mean(log(value+1)),
    mean.spot.count=mean(value[value>0])) 


spesum <- rename(spesum, Month='month(datetime)')
```

### Spatial distribution

This shows the number of stations where the class is spotted at least once. 




```{r}
ggplot(spesum,aes(x = Month, y= spotstation/20*100))+geom_bar(stat="identity", position=position_dodge())+facet_wrap(~variable)+ylab('Number of stations where spotted (%)')+
  scale_x_continuous(breaks = 1:12, labels = month.abb)+
  theme(axis.text.x = element_text(angle = 60, vjust = 0.5, hjust=1))
```

## Applying cluster stuffs

Requires transformation back to wide format. 
```{r}
d.spe <- spedf %>% 
  subset(select=c("datetime", "Sites", "variable","value")) %>%
  dcast(datetime + Sites ~ variable)


casteddates <- d.spe$datetime
castedSites <- d.spe$Sites

d.spe <- d.spe %>%
  select(-c("Sites","datetime")) 

d.spe
```

```{r}

# Bray-Curtis Dissimilarity for raw data
spe.db <- vegdist(d.spe)

# Bray-Curtis Dissimilarity for Log-transformed data
spe.log <- decostand(d.spe,'log')
spe.dbL <- vegdist(spe.log)

# Chord distance
spe.norm <- decostand(d.spe,'norm')
spe.dc <- dist(spe.norm)

# Hellinger
spe.hel <- decostand(d.spe,'hel')
spe.dh <- dist(spe.hel)

coldiss(spe.dc,byrank = FALSE,diag=FALSE); title("Chord")
coldiss(spe.dh,byrank = FALSE,diag=FALSE); title("Hellinger")

spe.norm <- decostand(d.spe,"norm")
spe.dist <- dist(spe.norm)
```



```{r}
spe.dist.single   <- hclust(spe.dist, method= "single")
plot(spe.dist.single)

spe.dist.complete <- hclust(spe.dist, method= "complete")
plot(spe.dist.complete)

spe.dist.average  <- hclust(spe.dist, method= "average")
plot(spe.dist.average)

spe.dist.centroid <- hclust(spe.dist, method= "centroid")
plot(spe.dist.centroid)

spe.dist.ward     <- hclust(spe.dist, method= "ward.D2")
spe.dist.ward$height <- sqrt(spe.dist.ward$height)
plot(spe.dist.ward)

```
"Average" gives the best cophenetic correlation. 

```{r}

cophDF <- data.frame(method = c("Single","Complete","Average","Centroid","Ward"), CophCorr= NA)
rownames(cophDF)<- cophDF$method

spe.dist.single.coph <- cophenetic(spe.dist.single)
cophDF["Single","CophCorr"] <- cor(spe.dist,spe.dist.single.coph)

spe.dist.complete.coph <- cophenetic(spe.dist.complete)
cophDF["Complete","CophCorr"] <- cor(spe.dist,spe.dist.complete.coph)

spe.dist.average.coph <- cophenetic(spe.dist.average)
cophDF["Average","CophCorr"] <- cor(spe.dist,spe.dist.average.coph)

spe.dist.centroid.coph <- cophenetic(spe.dist.centroid)
cophDF["Centroid","CophCorr"] <- cor(spe.dist,spe.dist.centroid.coph)

spe.dist.ward.coph <- cophenetic(spe.dist.ward)
cophDF["Ward","CophCorr"] <- cor(spe.dist,spe.dist.ward.coph)

cophDF

```

```{r}

finalClust <- spe.dist.average

plot(finalClust$height, nrow(d.spe):2, type="S",
     main = "Fusion levels - Chord - Average (UPGMA)", 
     ylab = "k (number of clusters)", 
     xlab = "h (node height)", 
     col="grey")
text (finalClust$height, nrow(d.spe):2,nrow(d.spe):2,col="red", cex=0.8)
```

```{r}
kfix <-6

spebc.single.g    <- cutree(spe.dist.single,k=kfix)
spebc.complete.g  <- cutree(spe.dist.complete,k=kfix)
spebc.average.g   <- cutree(spe.dist.average,k=kfix)
spebc.centroid.g   <- cutree(spe.dist.centroid,k=kfix)
spebc.ward.g      <- cutree(spe.dist.ward,k=kfix)


table(spebc.average.g, spebc.centroid.g)

```

```{r}

```


```{r}
asw <- numeric(nrow(d.spe))


for (k in 2:(nrow(d.spe)-1)){
  sil <- silhouette(cutree(finalClust, k=k), spe.dist)
  asw[k]<-summary(sil)$avg.width
}

k.best <- which.max(asw)

plot( 1:nrow(d.spe) , asw, type= 'h', 
      main = paste0("Silhouette optimal number of cluster for ", finalClust$method,',',finalClust$dist.method))
axis(1, k.best, col='red')
points(k.best, max(asw), col='red')
```
```{r}

finalk     <- 6

cutg <- cutree(finalClust,k=finalk)
sil  <- silhouette(cutg, spe.dist) 
silo <- sortSilhouette(sil)
rownames(silo) <- row.names(d.spe)[attr(silo,"iOrd")]

cutgo <- cutg[attr(silo,"iOrd")]

plot(silo, main= paste0( "silhouette Plot - ", finalClust$method,',',finalClust$dist.method," - k = ", finalk),
     cex.names=0.8, col=cutgo+1, nmax.lab = 100)
```

```{r}
spe.chwo<-reorder.hclust(finalClust,spe.dist)
hcoplot(finalClust, spe.dist, k=finalk)

```


```{r}
d.spe.clust       <- decostand(d.spe,'total')
d.spe.clust$datetime <- casteddates
d.spe.clust$g     <- cutg
d.spe.clust$sil   <- sil[,3]

```


```{r Visu Spe 2}
ggplot(melt(d.spe.clust, id.vars = c("datetime","g","sil")), aes(x=g, y=(value), group=factor(g))) +
  geom_boxplot()+
    geom_point(aes(color=sil))+scale_color_viridis_c()+
    facet_wrap(~variable)+theme_bw()


ggplot(melt(d.spe.clust, id.vars = c("datetime","g","sil")), aes(x=variable, y=(value), group=variable)) +
  geom_boxplot()+
    geom_point(aes(color=sil))+scale_color_viridis_c()+
    facet_wrap(~g)+theme_bw()+ theme(axis.text.x = element_text(angle = 60, vjust = 0.5, hjust=.4))

```


```{r}
d.spe.clust %>% filter(g==3) #%>% select(c('Sites','Mois'))
```


```{r, fig.height=5}
d.spe.clust$month<-factor(month(d.spe.clust$datetime))
d.spe.clust$year<-factor(year(d.spe.clust$datetime))

drel <- d.spe.clust %>% group_by(month) %>% summarize ( rel.1= sum(g==1)/length(g),
                         rel.2= sum(g==2)/length(g),
                         rel.3= sum(g==3)/length(g),
                         rel.4= sum(g==4)/length(g),
                         rel.5= sum(g==5)/length(g),
                         rel.6= sum(g==6)/length(g))

ggplot(melt(drel,id.vars = c('month')), aes(x=month, y=value, fill=variable)) +
  geom_bar(stat = 'identity')+
  scale_x_discrete(breaks=seq(1,12),labels=month.abb, name="")+theme_bw()
```

```{r, fig.width=12}

d.spe.clust$Sites <- castedSites

d.spe.clust$Mois <- as.numeric(month(d.spe.clust$datetime))

bibi <- merge(d.spe.clust, select(df,c("Sites",'X',"Y")), by = "Sites")


ggmap(bmap)+
    geom_point(bibi,  mapping = aes(x = X, y = Y,color=factor(g)), size=5)+
    facet_wrap('Mois',labeller=as_labeller(monthnames))+
    scale_color_brewer(palette = 'Set3')


```


```{r}
spe.T<-t(d.spe)

spe.t.chi <- decostand(spe.T, method = "chi.square")
spe.t.D16 <- dist(spe.t.chi)

coldiss(spe.t.D16,diag = TRUE)

```


```{r, fig.width=20}

require(GGally)

d.env <- envdf %>% 
  subset(select=c("datetime", "Sites", "variable","value")) %>%
  dcast(datetime + Sites ~ variable)%>%
  select(-c("Sites","datetime")) 

ggpairs(d.env, columns = 1:16)


# , lower = panel.smooth, 
#       upper = panel.cor, diag = panel.hist, main= "Kendall Correlation Matrix", method="kendall")
# 
# pairs(d.env.allvalid, lower.panel = panel.smooth, 
#       upper.panel = panel.cor, diag.panel = panel.hist, main= "Pearson Correlation Matrix", method="pearson")
# 
# pairs(d.env.allvalid, lower.panel = panel.smooth, 
#       upper.panel = panel.cor, diag.panel = panel.hist, main= "Spearman Correlation Matrix", method="spearman")

```

## Ordination

```{r, fig.width=10}
envdf

ggplot(envdf, aes(x=value))+
         geom_histogram(bins=50)+
         geom_density()+
         facet_wrap(~variable, scales = 'free')+
  theme_bw()
```

```{r}
d.env <- envdf %>% filter(month(datetime)!=6)%>%
  subset(select=c("datetime", "Sites", "variable","value")) %>%
  dcast(datetime + Sites ~ variable)

casteddates <- d.env$datetime
castedSites <- d.env$Sites

d.env <- d.env %>%
  select(-c("Sites","datetime")) 

d.env
```




```{r}
require(tidyr)

transformations <- list(
  original = function(x) x,
  log = function(x) ifelse(x > 0, log(x), NA),
  sqrt = function(x) ifelse(x >= 0, sqrt(x), NA),
  standard = function(x) scale(x),
  stand_sqrt = function(x) scale(sqrt(x))
)


# Initialize an empty list to store the results
shapiro_results <- list()

# Loop over each transformation
for (transformation_name in names(transformations)) {
  transformation_func <- transformations[[transformation_name]]
  
  # Apply the transformation and store the results of the Shapiro-Wilk test
  shapiro_result <- d.env %>%
    mutate(across(everything(), transformation_func)) %>%
    summarise(across(everything(), ~shapiro.test(.x)$p.value))
  
  # Add the transformation name as a column
  shapiro_result <- shapiro_result %>%
    mutate(transformation = transformation_name)
  
  # Append the result to the list
  shapiro_results[[transformation_name]] <- shapiro_result
}

# Combine all the results into a single data frame
final_results <- bind_rows(shapiro_results) %>%
  pivot_longer(-transformation, names_to = "variable", values_to = "shapiro_p_value")

# Print the final results
print(final_results)

ggplot(final_results, aes(y=log10(shapiro_p_value), x=variable, color=transformation))+
         geom_point()+geom_hline(yintercept = -2)
       
```


```{r}

subset(dfvar, Type=='env')$Parameters

```


```{r}
d.env.sc<-decostand(sqrt(d.env), 'standardize')

env.pca <- rda(d.env.sc) # Argument scale=TRUE calls for a
                                  # standardization of the variables
env.pca
summary(env.pca) # Default scaling 2
summary(env.pca, scaling=1)

```


## Examine and plot partial results from PCA output
```{r, fig.height=5}

# Eigenvalues
ev <- env.pca$CA$eig

# Apply Kaiser-Guttman criterion to select axes
ev[ev > mean(ev)]

# Broken stick model
n <- length(ev)
bsm <- data.frame(j=seq(1:n), p=0)
bsm$p[1] <- 1/n
for (i in 2:n) {
	bsm$p[i] = bsm$p[i-1] + (1/(n + 1 - i))
}
bsm$p <- 100*bsm$p/n
bsm

# Plot eigenvalues and % of variance for each axis
par(mfrow=c(2,1))
barplot(ev, main="Eigenvalues", col="bisque", las=2)
abline(h=mean(ev), col="red")	# average eigenvalue
legend("topright", "Average eigenvalue", lwd=1, col=2, bty="n")
barplot(t(cbind(100*ev/sum(ev),bsm$p[n:1])), beside=TRUE, 
	main="% variance", col=c("bisque",2), las=2)
legend("topright", c("% eigenvalue", "Broken stick model"), 
	pch=15, col=c("bisque",2), bty="n")
```

```{r, fig.height=4}	
# Same plots using a single function:
# Plot eigenvalues and % of variance for each axis
evplot(ev)
```

```{r}
plot(env.pca)
```

```{r, fig.height=4}
# Plots using cleanplot.pca
# A rectangular graphic window is needed for the two plots
#source('../Functions/cleanplot.pca.R')
par(mfrow=c(1,2))
cleanplot.pca(env.pca, ax1=1, ax2=2, scaling=1, plot.sites=TRUE, 
             plot.spe=TRUE, label.sites=FALSE, label.spe=TRUE, cex.char1=0.7)
cleanplot.pca(env.pca, ax1=1, ax2=2, scaling=2, plot.sites=TRUE, 
             plot.spe=TRUE, label.sites=FALSE, label.spe=TRUE, cex.char1=0.7)

```
```{r}
# Perform PCA
pca_result <- rda(d.env.sc)

# Summary of PCA result
summary(pca_result)

# Extract PCA scores for plotting
scores <- scores(pca_result, display = "sites")
scores_df <- as.data.frame(scores)
#scores_df$SampleID <- df$SampleID

# Plot PCA
p <- ggplot(scores_df, aes(x = PC1, y = PC2)) +
  geom_point() +
  #geom_text(aes(label = SampleID), vjust = -1) +
  labs(title = "PCA of Environmental Data",
       x = "PC1",
       y = "PC2") +
  theme_minimal()

# Display plot
print(p)

# PCA loadings
loadings <- scores(pca_result, display = "species")
loadings_df <- as.data.frame(loadings)

# Print loadings
print(loadings_df)
```



```{r}
# Get the site scores, scaling 1
sit.sc1 <- scores(env.pca, display="wa", scaling=1)

d.env$PCA1 <- sit.sc1[,1]
d.env$PCA2 <- sit.sc1[,2]
d.env$wdate <- casteddates

ggplot(d.env, aes(x=PCA1, y=PCA2, color=log(Chla)))+
  geom_point()+scale_color_viridis_c()+
  geom_vline(xintercept = 0)+
  geom_hline(yintercept = 0)+
theme_bw()
```



```{r}
ggplot(d.env, aes(x=PCA1, y=log(Chla)))+
  geom_point()+
theme_bw()

ggplot(d.env, aes(x=wdate, y=PCA1))+
  geom_point()+
theme_bw()

ggplot(d.env, aes(x=Temp, y=PCA1))+
  geom_point()+
theme_bw()

ggplot(d.env, aes(x=NO3, y=PCA1))+
  geom_point()+
theme_bw()

ggplot(d.env, aes(x=NO3, y=Temp, color=PCA1))+
  geom_point()+scale_color_viridis_c()+
theme_bw()
```





## Cyanobactéries & Oxygène 


That's a good one but the colors should be replaced by cluster of stations. 
```{r}
ddf <- df %>% filter( !is.na(OD) & !is.na(CYAN) & CYAN!=0 &month(datetime)>3 & month(datetime)<13) # & CYAN!=0 )

ggplot(ddf, aes(x=OD, y = log10(CYAN+1), color = Sites))+
  #geom_smooth(method = 'lm')+
  geom_boxplot(aes(group=cut(OD, 5)), notch = FALSE)+theme_bw()+
    #geom_boxplot(aes(group=ntile(OD, 5)), notch = FALSE)+theme_bw()+
  geom_point()+xlab('Dissolved Oxygen - [mg/l]')+ylab('Cyanobacteries: Abundance - [log(ind/ml)]')



ggplot(ddf, aes(x=OD, y = CYAN, color = Sites))+
  #geom_smooth(method = 'lm')+
  geom_boxplot(aes(group=cut(OD, 5)), notch = FALSE)+theme_bw()+
    #geom_boxplot(aes(group=ntile(OD, 5)), notch = FALSE)+theme_bw()+
  geom_point()+xlab('Dissolved Oxygen - [mg/l]')+ylab('Cyanobacteries: Abundance - [ind/ml]')+scale_y_log10()


cor.test(ddf$OD, log(ddf$CYAN+1))
```
```{r}
df
```

